---
title: "dplyr for beginners"
author: "Earl F Glynn<br>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    theme: cerulean
    toc: yes
    toc_depth:  3
    toc_float:
      collapsed:  yes
      smooth_scroll: yes
    number_sections: yes
    code_folding:  show
---

```{r HmiscSetup, echo = FALSE}
# http://biostat.mc.vanderbilt.edu/wiki/Main/KnitrHtmlTemplate
require(Hmisc)    # provides knitrSet and other functions
knitrSet(lang = 'markdown',   # If using blogdown: knitrSet(lang='blogdown')
         fig.align = 'left',
         w = 6.5,
         h = 4.5,
         cache = FALSE)
```

`r hidingTOC(buttonLabel = "Outline")`

Several `palmerpenguins` examples are based on Rebecca Barter's 
[Across (dplyr 1.0.0): applying dplyr functions simultaneously across multiple columns](http://www.rebeccabarter.com/blog/2020-07-09-across/)

```{r setup2, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment    = NA)

time.1 <- Sys.time()
```

# Setup {.tabset .tabset-fade .tabset-pills}

##  {.active}

## Packages                                       

```{r}
library(tidyverse)
library(kableExtra)  

library(readxl)

library(palmerpenguins)
```

## Helper function

```{r}
Show <- function(data, caption = "", bigMark = "", ...)
{
  data                                            %>%
  kable("html", caption = caption,
        format.args = list(big.mark = bigMark))   %>%
  kable_styling(bootstrap_options = c("striped", "bordered", "condensed"),
                position = "left",
                full_width = FALSE, ...)
}
```
            
# Overview of Palmer Penguins

## `str`

```{r}
str(penguins)        # `str` from utils package
```

## `glimpse`

```{r}
glimpse(penguins)    # `glimpse` from `tibble` package
```

# Slices

## `slice_head` 

Selected first and last rows

```{r}
penguins   %>%    # `%>%` is `magrittr` pipe operator; read "then"
  slice_head(n = 3) 
```

```{r}
penguins             %>%    
  slice_head(n = 3)  %>%
  Show()          # see section 1.3 "Helper function"
```

```{r}
penguins      %>%    
  slice(1:3)  %>%
  Show()          
```

```{r}
penguins                     %>%    # Use this with databases
  filter(row_number() <= 3)  %>%
  Show()          
```


## `slice_tail`

```{r}
penguins             %>%
  slice_tail(n = 3)  %>%
  Show()
```

```{r}
penguins                           %>%    # Use this with databases
  filter(row_number() >= n() - 2)  %>%
  Show()          
```

## Largest mass

### `top_n`

```{r}
penguins                %>% 
  top_n(3, body_mass_g) %>%    # name `top_n` is superseded by `slice_max`
  Show()
```

### `slice_max`

```{r}
penguins                        %>% 
  slice_max(body_mass_g, n = 3) %>%   
  Show()
```


# Frequency Counts 

## "Old" way

### 1D

```{r}
table(penguins$species, useNA = "ifany")
```

### 2D

```{r}
table(penguins$species, penguins$island, useNA = "ifany")
```

## "New" Way

### 1D

```{r}
penguins         %>%
  count(species) %>%     
  Show()
```

More general approach

```{r}
penguins             %>%
  group_by(species)  %>%
  summarise(n = n(),
            .groups = "drop") %>%  
  ungroup()          %>%
  Show()
```

### 2D

"long" format

```{r}
penguins                      %>%
  group_by(species, island)   %>%
  summarise(n = n(),                  # sometimes must spell as `summarise`
            .groups = "drop") %>%
  ungroup()                   %>%
  Show()
```

"wide" format 

```{r}
penguins                             %>%
  group_by(species, island)          %>%
  summarise(n = n(),                  
            .groups = "drop")        %>%
  ungroup()                          %>%
  pivot_wider(names_from  = island,             # `tidyr` function
              values_from = n,
              values_fill = list(n = 0))  %>%
  Show()
```

`spread` is a `retired` function but often works "better" IMHO than `pivot_wider`

# Gentoo Subset

```{r}
gentoo <-
  penguins                     %>%
  filter(species == "Gentoo")  %>%
  select(starts_with("bill_"), sex, year)   # options:  `ends_width`, `contains`

dim(gentoo)
```

```{r}
head(gentoo, 3) %>% Show()
```

## Summaries

### Longer

```{r}
gentooBySexYear <-
  gentoo               %>%
  group_by(sex, year)  %>%
  summarise(n = n(),
            across(c(bill_length_mm, bill_depth_mm), 
                   mean, .names = "mean_{col}"),
            .groups = "drop")         %>%
  ungroup()                           %>%
  rename_with(function(x){str_replace_all(x, "mean_bill_|_mm", "")},
         starts_with("mean_bill_"))   

gentooBySexYear %>%
  Show(caption = "Mean Bill Length and Bill Depth [mm]")
```

## Wider

```{r}
gentooBySexYear        %>%
  filter(!is.na(sex))  %>%
  pivot_wider(
    names_from  = year, 
    values_from = c(n, length, depth)
  )                    %>%
  Show(caption = "Mean Bill Length and Depth [mm]")
```

# Palmer Penguin Summaries

## Global

```{r}
penguins                   %>%
  summarise(
             n             = n(),
             
             nSpecies      = n_distinct(species),
             nameSpecies   = str_flatten(species %>% unique() %>% sort(), 
                                         collapse = "|"),
             
             nIsland       = n_distinct(island),
             nameIsland    = str_flatten(island %>% unique() %>% sort(), 
                                         collapse = "|"),
             
             nSex          = n_distinct(sex),
             nameSex       = str_flatten(sex %>% unique() %>% sort(), 
                                         collapse = "|"),
             
             nYear    = n_distinct(year),
             minYear  = min(year, na.rm = TRUE),
             maxYear  = max(year, na.rm = TRUE)
           )          %>%
  Show()
```

## With Across

```{r}
flatString <- function(variable)
{
  str_flatten({{variable}}  %>%  unique()  %>%  sort(),
              collapse ="|")  
}
```

Factor with NA to string "-missing-"

```{r}
penguins              %>%
  mutate(sex = as.character(sex))             %>%  # force factor to character string
  mutate(sex = replace_na(sex, "-missing-"))  %>%
 
  summarise(
             n        = n(),
             across(c(species, island, sex, year), n_distinct, .names = "n_{col}"),
             across(c(species, island, sex, year), flatString, .names = "names_{col}")
           )          %>%
  Show()
```


## Counts of missing values

```{r}
penguins                             %>%
  summarise(across(everything(), 
                   ~sum(is.na(.))))  %>%
  Show()
```

# Recoding

```{r}
penguinRecoded <- 
  penguins                        %>%
  mutate(species = 
           recode(species,
                  "Adelie"    = "Penguin1",
                  "Chinstrap" = "Penquin2",
                  "Gentoo"    = "Penguin3"),
         
         island =
           recode(island,
                  "Biscoe"    = "Island1",
                  "Dream"     = "Island2",
                  "Torgersen" = "Island3"),
         
         sex = replace_na(as.character(sex), "-missing")
        )
```

```{r}
glimpse(penguinRecoded)
```

# Joins

## Add "info" 

Add `dimension` info to `fact` table

```{r}
info <- read_xlsx("infoPenguins.xlsx")

info %>% Show()
```

```{r}
infoPenguins <-
  penguins  %>%
  inner_join(info, by = "species")

glimpse(infoPenguins)
```

## Control selections

Use "tagging" especially with large lists for selections without much typing.

```{r}
target <- 
  read_xlsx("targetPenguins.xlsx")  %>%
  filter(tag == "x")                %>%  # Can type "x" reliably
  select(-tag)

target %>% Show()
```

```{r}
targetPenguins <-
  penguins  %>%
  inner_join(target, by = "species")

glimpse(targetPenguins)
```

```{r}
targetPenguins  %>%  count(species)  %>%  Show()
```

Use `filter` instead of `inner_join`

```{r}
targetPenguins2 <-
  penguins  %>%
  filter(species %in% target$species)

glimpse(targetPenguins2)
```

# Quantiles

See [dplyr 1.0.0: new summarise() features](https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-summarise/#quantiles).

```{r}
Qs <- c(0, 0.10, 0.25, 0.50, 0.75, 0.90, 1.00)
```

## `bill_length all`

```{r}
quantile(penguins$bill_length_mm, Qs, na.rm = TRUE)
```

```{r}
penguins       %>%
  summarise(quantiles   = Qs,
            bill_length = quantile(bill_length_mm, Qs, na.rm = TRUE))  %>%
  Show()
```


## `bill_length` by `sex`

```{r}
penguinQuantiles <- 
  penguins       %>%
  group_by(sex)  %>%
  summarise(quantiles   = paste0("Q", 100*Qs),
            bill_length = quantile(bill_length_mm, Qs, na.rm = TRUE),
            .groups = "drop")  %>%
  ungroup()                    %>%
  pivot_wider(
    names_from  = quantiles,
    values_from = bill_length
  )

penguinQuantiles  %>%
  Show()
```

# Other Useful Notes

* [dplyr 1.0.0: working across columns](https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-colwise/).

* [dplyr 1.0.0: select, rename, relocate](https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-select-rename-relocate/).

* Dario Radečić's [How to Analyze Data with R: A Complete Beginner Guide to dplyr](https://appsilon.com/r-dplyr-tutorial/).

* Rasmus Bååth's [The Tidyverse in a Table](http://www.sumsar.net/blog/2020/12/tidyverse-in-a-table/).

* Emily Riederer's [Generating SQL with {dbplyr} and sqlfluff](https://emilyriederer.netlify.app/post/sql-generation/).

* HighlandR's [Solving small data problems with data.table](https://johnmackintosh.com/2020-08-11-short-problems/).

* [Tidyverse Tips](https://www.r-bloggers.com/2020/11/tidyverse-tips/).

* [Understanding Non-Standard Evaluation](https://thomasadventure.blog/posts/understanding-nse-part1/).

# Fini {.tabset .tabset-fade .tabset-pills}

## {.active}

## Session Info

```{r}
devtools::session_info()
```

</div>

```{r, echo=FALSE}
time.2 <- Sys.time()
processingTime <- paste("Processing time:",
                        sprintf("%.1f",
                                as.numeric(difftime(time.2, time.1,
                                                    units="secs"))), "secs\n")
```

`r processingTime`

`r format(Sys.time(), "%Y-%m-%d %H:%M")`                                                                                      